---
title: "lab 3"
author: "John Barmen"
date: "3/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part 1
```{r}
#install.packages('readxl')
library(ggplot2)
library(dplyr)
library(stringr)
dat <- readxl::read_xls("/Users/jrdba/OneDrive/Documents/DS202/ds202_lab3/GSS.xls")
dat <- na.omit(dat)
names(dat)
dim(dat)
names(dat)<-str_replace_all(names(dat), c(" " = "." , "," = "" ))
dat<- rename(dat,
    year = Gss.year.for.this.respondent,
    relig = Rs.religious.preference,
    income = Total.family.income,
    edu = Highest.year.of.school.completed,
    MarStat = Marital.status,
    bal = Ballot.used.for.interview,
    happiness = General.happiness,
    polParty = Political.party.affiliation,
    sex = Respondents.sex,
    age = Age.of.respondent,
    id = Respondent.id.number
  )
names(dat)
dim(dat)
```

### Part 2
1)
```{r 1}
dat$edu[dat$edu=="No answer"| dat$edu=="Don't know"] <- -1
dat$edu<-strtoi(dat$edu)
avg<-dat %>%
  group_by(MarStat) %>%
  summarize(Average=mean(edu)) %>%
  ungroup()%>%arrange(-Average) %>%
  data.frame
avg

```
2)
```{r 2}
smort <- subset(dat, ((MarStat == "Never married" & edu > 13.35210)|
                        (MarStat == "Divorced" & edu > 13.07889)|
                        (MarStat == "No answer" & edu > 13.03704)|
                        (MarStat == "Married" & edu > 12.93718)|
                        (MarStat == "Separated" & edu > 11.80419)|
                        (MarStat == "Widowed" & edu > 11.12742)))
str(smort)
```

3)
```{r 3}

dat$MarStat <-factor(dat$MarStat)
ggplot(dat, aes(happiness, MarStat))
dat$happiness[dat$happiness == "No answer"| dat$happiness == "Don't know"|dat$happiness == "Not applicable"] <- 0
dat$happiness[dat$happiness == "Not too happy"] <- 0
dat$happiness[dat$happiness == "Pretty happy"] <- 1
dat$happiness[dat$happiness == "Very happy"] <- 1
dat$happiness <- strtoi(dat$happiness)
ggplot(data=dat, aes(x=happiness))+geom_histogram()+facet_wrap(~MarStat)
temp <- data.frame(dat$happiness, dat$MarStat)
summary(temp)
```
To better manipulate the data I corresponded various responses to levels of happiness, 0 for not happy, 1 for happy. by far married people are more happy compared to their unhappy counterparts, but unmarried people are also fairly happy.  

4)
```{r 4}
dat$polParty <-factor(dat$polParty)
ggplot(data=dat, aes(x=polParty))+geom_bar()+facet_wrap(MarStat~happiness)
```

5)
```{r 5}
ggplot(data=dat, aes(y=edu, x=MarStat, fill=MarStat))+geom_boxplot()
```
Based on averages the No answer group of marige status were the best educated, otherwise ther are fairly even in two groups, Divorced/Married/Never married are about the same being consistantly higher than the Seperated/Widdowed which are fairly even with each other.
6)
```{r 6}
dat$age[dat$age == "No answer"| dat$age == "Don't know"] <- -1
dat$age[dat$age == "89 or older"] <- 89
dat$age <- strtoi(dat$age)
ggplot(data=dat, aes(y=age, x=MarStat, fill=MarStat))+geom_boxplot()
```

```{r}
ggplot(data=dat, aes(x=happiness))+geom_histogram()+facet_wrap(~relig)
```















